Pre-Trained Denoising Autoencoders Long Short-Term Memory Networks as probabilistic Models for Estimation of Distribution Genetic Programming

### 2023-01-01 paper.Rmd
autogit

The DAE-GP algorithm first described by [@dae-gp_2020_rtree] trains a DAE-LSTMs for each generation $g$ of the search from scratch.
@dae-gp_2022_symreg and @daegp_explore_exploit suggests the incorporation of a pre-training strategy into the evolutionary search as a possible way of improving the model quality of DAE-GP.
The key idea is to pre-train an initial DAE-LSTM on a large population of candidate solutions and to use the weights of this initial model in each generation of the search as a starting point.
This thesis studies the influence of using pre-trained DAE-LSTMs in DAE-GP for symbolic regression, especially looking at the influence in generalization behavior and the overall quality of solutions found by DAE-GP. 
The aim of this study is to answer the question if a pre-training strategy can help to improve generalization in DAE-GP and if so, analyze how this improved generalization ability benefits the overall performance of DAE-GP.

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM model $M_g$ that are used in each DAE-GP generation $g$ for $g\in{1,...,g_{max}}$. 
Instead of initializing and optimizing each model from scratch as done in previous work (e.g. [@dae-gp_2020_rtree] [@dae-gp_2022_symreg]), a separate DAE-LSTM network $\hat{M}$ will be trained on a large population of randomly initialized solutions.  
The optimal parameters obtained after finishing the training procedure of $\hat{M}$ are then used as the starting parameters for each following DAE-LSTM model $M_g$ for $g\in{1,...,g_{max}}$.

@@

The probably most obvious motivation for using a pre-training strategy in DAE-GP is based on the fact, that initializing $M_g$ with pre-trained weights from $\hat{M}$ is likely to reduce the amount of training epochs that are necessary until the point of training error convergence is reached.

@@

The DAE-GP algorithm, first described by @dae-gp_2020_rtree, is summarized as a flowchart in figure 1. 
After setting the generation counter $g$ to $0$ and creating the initial population of solutions $P_0$, DAE-GP enters the main loop that checks if a termination criteria is satisfied (i.e. maximum number of generations or fitness evaluations is reached).
For each generation $g$ the fitness of all new individuals in $P_g$ is evaluated. During model building, DAE-GP selects a subset $X_g$ of the current population that is used as training data for the DAE-LSTM model $M_g$ which is trained to learn the properties of $X_g$. 
The trained model $M_g$ is then used for model sampling, partially mutated solutions from $P_G$ are propagated through $M_g$ to create a new population $P_{g+1}$. 
This process is repeated until the termination criteria is met and DAE-GP returns the highest fitness solutions.

@@

The pre-training strategy implemented for all experiments in this thesis is visualized as another flowchart in figure 2.
The main difference to regular DAE-GP is the inclusion of an initial pre-training phase where a separate population $\hat{P}$ is first initialized and then randomly split by half into $\hat{P_{train}}$ and $\hat{P_{test}}$. 
A DAE-LSTM model $\hat{M}$ is then trained to learn the properties of $\hat{P_{train}}$ using an early stopping training mode where we stop training as soon as the validation error for $\hat{P_{test}}$ converges. 
After the training for $\hat{M}$ is stopped, we freeze the current state of $\hat{M}$ and terminate the pre-training phase to start the DAE-GP phase. 
During the DAE-GP phase, the only difference to the traditional DAE-GP algorithm described in figure 1 is during model building: We initialize $M_g$ with the weights and biases of $\hat{M}$ before training it to learn the properties of $X_g$.

@@

>>>>

The DAE-GP algorithm first described by [@dae-gp_2020_rtree] trains a DAE-LSTMs for each generation $g$ of the search from scratch.
@dae-gp_2022_symreg and @daegp_explore_exploit suggests the incorporation of a pre-training strategy into the evolutionary search as a possible way of improving the model quality of DAE-GP.
The key idea is to pre-train an initial DAE-LSTM on a large population of candidate solutions and to use the pre-trained parameters of this initial model in each generation of the search as starting parameters for the current generation DAE-LSTM.
This thesis studies the influence of using pre-trained DAE-LSTMs in DAE-GP for symbolic regression, especially looking at the influence in generalization behavior and the overall quality of solutions found by DAE-GP. 
The aim of this study is to answer the question if a pre-training strategy can help to improve generalization in DAE-GP and if so, analyze how this improved generalization ability benefits the overall performance of DAE-GP.

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM model $M_g$ that are used in each DAE-GP generation $g$ for $g\in{1,...,g_{max}}$. 
Instead of initializing and optimizing each model from scratch as done in previous work (e.g. [@dae-gp_2020_rtree] [@dae-gp_2022_symreg]), a separate DAE-LSTM network $\hat{M}$ will be trained on a large population of randomly initialized solutions.  
The trainable parameters $\theta_{\hat{M}}$ obtained after finishing the training procedure of $\hat{M}$ are then used as the starting parameters for each following DAE-LSTM model $M_g$ for $g\in{1,...,g_{max}}$.

@@

The probably most obvious motivation for using a pre-training strategy in DAE-GP is based on the fact, that initializing $M_g$ with pre-trained weights $\theta_{\hat{M}}$ is likely to reduce the amount of training epochs that are necessary until the point of training error convergence is reached.

@@

The DAE-GP algorithm, first described by @dae-gp_2020_rtree, is summarized as a flowchart in figure §. 
After setting the generation counter $g$ to $0$ and creating the initial population of solutions $P_0$, DAE-GP enters the main loop that checks if a termination criteria is satisfied (i.e. maximum number of generations or fitness evaluations is reached).
For each generation $g$ the fitness of all new individuals in $P_g$ is evaluated. During model building, DAE-GP selects a subset $X_g$ of the current population that is used as training data for the DAE-LSTM model $M_g$ which is trained to learn the properties of $X_g$. 
The trained model $M_g$ is then used for model sampling, partially mutated solutions from $P_G$ are propagated through $M_g$ to create a new population $P_{g+1}$. 
This process is repeated until the termination criteria is met and DAE-GP returns the highest fitness solutions.

@@

The pre-training strategy implemented for all experiments in this thesis is visualized as another flowchart in figure §.
The main difference to regular DAE-GP is the inclusion of an initial pre-training phase where a separate population $\hat{P}$ is first initialized and then randomly split by half into $\hat{P_{train}}$ and $\hat{P_{test}}$. 
A DAE-LSTM model $\hat{M}$ is then trained to learn the properties of $\hat{P_{train}}$ using an early stopping training mode where we stop training as soon as the validation error for $\hat{P_{test}}$ converges. 
After the training for $\hat{M}$ is stopped, the current state of $\hat{M}$ is frozen and the optimized trainable parameters $\theta_{\hat{M}}$ are saved before terminating the pre-training phase to start the DAE-GP phase. 
During the DAE-GP phase, the only difference to the traditional DAE-GP algorithm described in figure § is during model building: We initialize $M_g$ with $\theta_{\hat{M}}$ before training it to learn the properties of $X_g$.

@@

<!-- Results for Energy Cooling Dataset  -->

```{bash compile_energyCooling_2hl_maxIndSize_fullRun_30gens, include=FALSE}
bash /Users/rmn/masterThesis/eda-gp-2020/experiments/energyCooling_2hl_maxIndSize_fullRun_30gens/re_evaluate_energyCooling_2hl_maxIndSize_fullRun_30gens.sh
```

### 2023-01-01 slides.Rmd
autogit

  1. das Generalisierungsverhalten von DAE-GP?
  2. die Qualität der gefundenen Programme?
  3. die Populationsdiversität?
  4. das Laufzeitverhalten?

@@

>>>>

  1. das Generalisierungsverhalten von DAE-GP?
  2. die Qualität der gefundenen Programme (Fitness/Programmlänge)?
  3. die Populationsdiversität?
  4. das Laufzeitverhalten?

@@

* Metaheuristik basierend auf genetischer Programmierung (GP)
* Einsatz von künstlichen, neuronalen Netzen zur Optimierung des GP Suchverhaltens
* Aktueller Forschungsstand:
  1. 
  2.

## Estimation of Distribution Genetic Programming (EDA-GP)

* Ersetzt die genetische Operatoren Mutation und Rekombination durch den Einsatz von probabilistischen Modellen

2 Phase Ansatz:

  1. Model Building: 
  2. Model Sampling:
